### 2019-02-28 UnderstandabilityPaper/machinelearning.tex
Fleshed out automated theorem proving section

%  - Understandability in AI
%  - Theorem proving as example:
%   - Least informative

\documentclass{article}
\begin{document}
I meant to send this earlier, in case it's useful. These are just some
things that came to mind regarding "explainable" AI and machine
learning. Lots of rambling, but hopefully some insight into current
machine-learning thinking (as I understand it, as somewhat of an
outsider!).

>>>>

%  - Understandability in AI
%  - Theorem proving as example:
%   - Least informative is a yes/no/unknown answer
%   -

\documentclass{article}
\begin{document}
The problem of making machine-generated artefacts more understandable to human
users is quite clear in the field of automated theorem proving~CITE: the task of
deciding whether a given formal statement follows from a given set of premeses.
The least informative approaches, such as those based on the popular \emph{SLD
  resolution} algorithm~CITE, produce nothing more than a ``yes'', ``no'' or
``unknown'' response. Not only is this devoid of \emph{explanation}, but it also
hides the effects of any bugs; requiring the user to either trust the results,
or verify the implementation.

Both of these issues can be mitigated by having the system instead generate a
\emph{proof object}: a formal argument for \emph{why} a given statement follows
or does not follow. Once generated, a proof object's validity can be checked
without requiring any knowledge of how it was created, thus avoiding the need to
trust or verify the (potentially complicated) search and generation
procedures. The \emph{de Bruijn criterion}~CITE describes theorem provers which
produce proof objects that are trivial to check by independent \emph{proof
  checker} programs (which are themselves easily verified, due to their
simplicity); examples are Coq~CITE and Isabelle~CITE.

Proof objects are not a complete solution to understandability, since they can
still be quite inscrutable to human users. This often depends on how closely
the chosen formal system is able to encode the user's ideas: for example, the
formal proof of Kepler's Conjecture was performed using a system of Higher Order
Logic (HOL)~CITE whose proof objects (natural-deduction style derivations),
whilst tedious, are in principle understandable to a user experienced with both
the software and problem domain. The same cannot be said of the Boolean
Pythagorean Triples problem, a statement of Ramsey theory involving the
structure of the natural numbers. Rather than taking a high-level approach like
HOL, the authors of~CITE analysed sets $\{0..n\}$ for larger and larger $n$,
encoding these restrictions of the problem into the language of boolean
satisfiability (SAT), and finding that the problem is unsatisfiable for
$n= 7825$, and hence for the natural numbers as a whole. The proof object for
$n = 7825$ contains 200 terabytes (compressable to 68 gigabytes) of
propositional logic clauses, demonstrating the unsatisfiability of the encoded
problem instance. Not only is this far too much for any human to comprehend, but
the concepts used in the proof (boolean formulae) are several layers removed
from the actual problem statement (natural numbers, subsets and pythagorean
triples).

Whilst ``low level'' formalisms like SAT are less understandable or explanatory
for users, they are far more amenable to automation than more expressive logics.
Despite the proof for the Boolean Pythagorean Triples problem being many orders
of magnitude larger than that of the Kepler Conjecture, the latter is well
beyond the ability of today's automated theorem provers due to its encoding in
HOL. Instead, it took 22 collaborators 9 years just to formalise the proof
(Hales had previously produced a less formal proof, hundreds of pages long and
accompanied by unverified software).

I meant to send this earlier, in case it's useful. These are just some
things that came to mind regarding "explainable" AI and machine
learning. Lots of rambling, but hopefully some insight into current
machine-learning thinking (as I understand it, as somewhat of an
outsider!).
